Article Title

### 2018-03-07 doc/turtles_plosone.tex
ken and jim comments. sex ratio table

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{Ensemble approaches for estimating congruence between species delimitation and morphological variation: comparing taxonomic hypotheses for the Pacific Pond Turtle (\textit{Emys marmorata})} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Peter D Smits\textsuperscript{1*},
Kenneth D Angielczyk\textsuperscript{2},
Bryan L Stuart\textsuperscript{3},
James F Parham\textsuperscript{4},
\\
\bigskip
\textbf{1} Department of Integrative Biology, University of California -- Berkeley, Berkeley, California, USA.
\\
\textbf{2} Integrative Research Center, Field Museum of Natural History, Chicago, Illinois, USA.
\\
\textbf{3} Section of Research and Collections, North Carolina Museum of Natural Sciences, Raleigh, North Carolina, USA.
\\
\textbf{4} John D. Cooper Archaeological and Paleontological Center, Department of Geological Sciences, California State University -- Fullerton, Fullerton, California, USA.
\\
\bigskip

@@

\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
  We investigated the morphometric identification of cryptic species using machine learning approaches by examining their implications for a recently proposed cryptic turtle species (\textit{Emys pallida}). We collected landmark-based morphometric data from 532 adult \textit{E. marmorata/``E. pallida''} museum specimens. We assigned a classification to each specimen for six different binning schemes based on geographic occurrence data recorded in museum collection archives. We used an ensemble of supervised machine learning approaches to determine which classification hypothesis was best supported by the data. In addition, we applied the same approach to two clear-cut examples, one consisting of eight unambiguously distinct species closely related to \textit{E. marmorata}, and the other consisting of two subspecies of \textit{Trachemys scripta}. Our results indicate that there is no clear ``best'' grouping of \textit{E. marmorata/``E. pallida''} based on plastron shape. In contrast, the analyses of the clear-cut examples produced near perfect classifications, demonstrating that the methods can recover correct results when an appropriate signal exists. Explanations for the lack of grouping in \textit{E. marmorata} include the possibility that genetic differentiation is not associated with plastron shape variation below the species level and/or that local selective pressures (e.g., from hydrological regime) overwhelm morphological differentiation. A reconsideration of the methods used to delimit \textit{``E. pallida,''} the lack of barriers to gene flow, the strong evidence for widespread admixture between lineages, and the fact that plastron shape can be used to delineate other emydine species and sub-species suggest that its lack of diagnosability most likely reflects the non-distinctiveness of this proposed taxon.

@@

These considerations have sparked interest in whether geometric morphometric analyses can capture fine-scale variation that can be used for identifying cryptic species. This would make the task of identifying and maintaining endangered or conserved groups much easier and could contribute to improved classifications of extinct taxa and populations. Most such studies focus on using morphometrics to discover differences between taxa that were identified by other means \cite{Polly2003,Zelditch2004,Gaubert2005b,Gunduz2007,Polly2007a,Demandt2009,Markolf2013,Fruciano2016}. Additionally, there has been work on automated taxon identification and classification of taxa into groups \cite{Baylac2003,Dobigny2003,MacLeod2007,VandenBrink2011,Vitek2017}, as well as the development of models that combine genetic, phenotypic, and geographic data to infer evolutionary units of interest \cite{Guillot2012}.

Here, we investigate the morphometric identification of cryptic species using machine learning approaches. We use an ensemble learning approach where multiple methods are used in order to look for consensus among their results. We test our approach on three datasets: plastron shape of eight species of closely related turtles, plastron shape of two subspecies of a single turtle species, and plastron shape of the \textit{Emys marmorata} species complex. In particular, we ask whether it is possible to determine which among a set of classification hypotheses best aligns with the observed morphology, and examine the implications of our results for the \textit{E. marmorata} complex.

% past work on automatic taxon identification and older approaches to classifying taxa
% why use machine learning methods
\subsection*{Background and study system}
Machine learning is an extension of known statistical methodology that emphasizes predictive accuracy and generality often at the expense of the interpretability of individual parameters \cite{Hastie2009}. Basic statistical approaches are supplemented by randomization, sorting, and partitioning algorithms, along with the maximization or minimization of summary statistics, in order to best estimate a general model for all data, both sampled and unsampled \cite{Hastie2009}. Machine learning approaches have found use in medical research, epidemiology, economics, and automated identification of images such as handwritten zip codes \cite{Hastie2009}. % additional/more specific CITATIONS?

There are two major classes of machine learning method: unsupervised and supervised learning. Unsupervised learning methods are used with unlabeled data where the underlying structure is estimated; they are analogous to clustering and density estimation methods \cite{Kaufman1990}. Supervised learning methods are used with labeled data where the final output of data is known and the rules for going from input to output are inferred. These are analogous to classification and regression models \cite{Breiman1984,Hastie2009}. Our application of the supervised learning approaches used in this study illustrates only a sampling of the various methods available for fitting classification models. The specific methods used in this study were chosen because they are suited for cases with more two or more response classes.

@@

More recently, the phylogeography of \textit{E. marmorata} and the possibility of cryptic diversity was investigated using molecular data \cite{Spinks2005,Spinks2010,Spinks2014}. Based on mitochondrial DNA, Spinks and Shaffer \cite{Spinks2005} recognized four subclades within \textit{E. marmorata}, a northern clade, a San Joaquin Valley clade, a Santa Barbara clade, and a southern clade. Analyses with nuclear DNA \cite{Spinks2010} and single-nucleotide polymorphism (SNP) data suggest a primarily north--south division in \textit{E. marmorata}, although these datasets differed from that of mitocondrial-based results of Spinks and Shaffer \cite{Spinks2005} in the location of the break point \cite{Spinks2014}. All three studies discussed the potential taxonomic implications of their results, with Spinks et al. \cite{Spinks2014} going so far as to strongly advocate for the recognition of at least two species (\emph{E. marmorata} and \emph{E. pallida}), and a possible third based on populations in Baja California. However, they did not discuss in detail the morphological characters that would help to diagnose these species beyond those specified by Seeliger \cite{Seeliger1945}. Given that these characters are variable within the proposed species, and that Holland \cite{Holland1992} described shell shape variation that might be consistent with this taxonomy, a geometric morphometric analysis of shell shape might provide a reliable way to diagnose groups (whether species or subspecies) within \textit{E. marmorata}.

@@

We choose to analyze plastron shape for multiple reasons. First, it is very easy to collect geometric morphometric data on plastron shape from two-dimensional pictures as the structure is virtually flat. This approach allows both museum specimens and individuals in the field to be analyzed together. Second, previous work has suggested that there are strong differences in plastron shape among traditionally-recognized emydine species \cite{Angielczyk2007,Angielczyk2011,Angielczyk2013a}. Finally, due to these previous studies a large dataset was readily available.

@@

Three different geometric morphometric datasets describing turtle plastron variation were assembled for this analysis: 1) specimens from eight distinct emydine species; 2) \textit{T. scripta} specimens from the two main subspecies (\textit{T. scripta elegans} and \textit{T. scripta scripta}); and 3) \textit{E. marmorata} specimens from across the species' geographic range. The first two datasets are intended to serve as a test of whether machine learning techniques can differentiate species-level groupings of emydine turtles using plastron shape. We expect that the first case represents a low complexity dataset because of the high level of plastron shape disparity that exists among these species \cite{Claude2003a,Claude2006,Angielczyk2011}, whereas the second dataset should be relatively higher in complexity and more analogous to the \textit{E. marmorata} example. We predict that the \textit{E. marmorata} dataset should be of the highest complexity and our greatest challenge given the finding that only very subtle differences existed between geographically-distinct populations \cite{Holland1992}.
%We expect the first dataset represents a low-complexity dataset given the high level of plastron shape disparty that exists among these species \cite{Angielczyk2011}, whereas the second dataset should be relatively higher in complexity than the first. Ultimately, we expect that the third dataset to be of the highest complexity and our biggest analytical challenge. For all datasets we chose to focus on adults because significant changes in plastron shape occur over the course of ontogeny in \textit{E. marmorata} and other emydines \cite{Angielczyk2013a}.
% three turtle datasets
%   eight species
%   Trachemys species
%   Emys complex
% 26 landmarks

@@

The final dataset is of 532 adult \textit{E. marmorata} museum specimens, though not all specimens were able to be assigned a class for all schemes (Fig. \ref{fig:map}). These specimens represent a subset of those included in previous studies \cite{Angielczyk2007,Angielczyk2011,Angielczyk2013a}. Because previous studies did not use vouchered specimens \cite{Spinks2005,Spinks2010,Spinks2014} we were not able to directly sample the individuals in their studies. Instead, our specimen classifications were based solely on the geographic information and not explicit assignment using molecular data. For each taxonomic hypothesis, specimens were assigned to one of the possible classes based on geographic occurrence data recorded in museum collections. In cases where precise latitude and longitude information were not available we estimated them from other locality information. Because the exact barriers between different biogeographic regions are unknown and unclear, we represented each hypothesis with multiple possible realizations representing the classification uncertainty for specimens present at the geographic boundaries. The taxonomic hypotheses and sub-hypotheses for \textit{E. marmorata} used here are presented in Table \ref{tab:hypotheses} and Fig. \ref{fig:map}.

For our first three schemes \cite{Spinks2010} we used three binning schemes. All three schemes include a class for \textit{E. marmorata} specimens from northern populations (marm) as well as a class for those assigned to \textit{E. pallida} (pall) and an intergrade zone in the Central Coast Ranges (CCR). The schemes differ in the assignment of samples from the San Joaquin Valley (Fig. \ref{fig:map}). Scheme SP10.1 and SP10.2 differ in the assignment of specimens from the western San Joaquin Valley to either CCR or marm reflecting uncertainty regarding their genetic affinity as explained above. In scheme SP10.3 these specimens are assigned to a San Joaquin class reflecting the mitochondrial distinctiveness shown by Spinks and Shaffer \cite{Spinks2005}. For our second two schemes \cite{Spinks2014} we used two binning schemes with SP14.1 being based on their phylogenetic network analysis and SP14.2 being based on their Bayesian species delimitation analysis. The latter scheme requires the addition of two new classes, ``Baja'' and ``Foothill,'' to accommodate the genetic groupings recovered by the SNP Structure analysis that was used to create the guide tree for the BPP species delimitation analysis \cite{Spinks2014}. Finally, we proposed a conservative morphological hypothesis (``Morph'') in order to compare the molecular hypotheses with something approximating the original taxonomic hypothesis for the group; this scheme is made up solely of the marm and pall classes from the SP10.3 scheme.

@@

Sex was known only for a subset of the total dataset and was not included as a predictor of classification. Instead, we estimated the degree by which specimens cluster morphologically by sex in order to determine how much of a potential biasing factor sexual dimorphism could be for our analysis of the \textit{E. marmorata} species complex (see below).
%The scheme names are as follows: Mito 1 and 2 correspond to \cite{Spinks2005}, Mito 3 corresponds to \cite{Spinks2010}, Morph 1 and Morph 2 correspond to \cite{Holland1992}, and Nuclear corresponds to \cite{Spinks2014}. % CHANGE

Following previous work on plastron shape \cite{Angielczyk2007,Angielczyk2011,Angielczyk2013a}, we used TpsDig 2.04 \cite{Rohlf2005} to digitize 19 two-dimensional landmarks (Fig. \ref{fig:plastra}). Seventeen of the landmarks are at the endpoints or intersection of the keratinous plastral scutes that cover the plastron. Twelve of the landmarks were symmetrical across the axis of symmetry. Because damage prevented the digitization of all the symmetric landmarks in some specimens, we reflected landmarks across the axis of symmetry (i.e. midline) prior to analysis and used the average position of each symmetrical pair. In cases where damage or incompleteness prevented symmetric landmarks from being determined, we used only the single member of the pair. We conducted all subsequent analyses on the resulting ``half'' plastra. We superimposed the plastral landmark configurations using generalized Procrustes analysis \cite{Dryden1998a}, after which we calculated the principal components (PC) of shape using the \texttt{shapes} package for R \cite{R2016,Dryden2013}. All specimens were used for superimposition, after which the subset labeled for each of the schemes were used in model training and testing (see below).

\subsection*{Biasing effects}
% digitization
We estimated the possible effect of digitization error \cite{Arnqvist1998,Cramon2007,Munoz-MunozF.2010} on our results by comparing within-specimen (replicated) Procrustes distances to the distances between classification scheme centroids. Ten randomly-selected \textit{E. marmorata} specimens were each digitized four times, with the original set of digitized coordinates serving as a fifth replicate. These 50 landmark configurations were then Procrustes superimposed. A range of four Procrustes distances was then calculated as the average of the pairwise distances between each of the replicate configurations of a given specimen.

For each specimen, the difference in shape caused by digitization was calculated as the mean of all pairwise Procrustes distances between the five replicates of that specimen. The average distance between any two digitizations was calculated as the mean of all pairwise Procrustes distances between all replicates for all specimens. The ratio between these two values was used to assess the magnitude of variation caused by digitization. The goal of this ratio is to determine if the within group distances are on average smaller than the between individual distances; a value of 0 indicates perfect grouping, a value of 1 indicates no difference between grouping and no grouping, and a value of 1+ indicates that the grouping is counter-intuitive to the data.

\textit{Emys marmorata} is known to display sexual dimorphism in plastral shape, particularly the presence of a plastra concavity in males \cite{Seeliger1945}. To test for biases resulting from sexual dimorphism in our \textit{E. marmorata} dataset, we used a simple permutation test to determine if the distance between the mean female and male shapes is greater than expected when the sex labels are randomly shuffled. Because not all of our specimens have sex identifications associated with them, this analysis was done using a subset of the data (257 of 532). This analysis was also done for each of the classification schemes in order to determine if sex is a biasing factor within these groups. As with the total dataset, only a limited number of specimens in each classification scheme have sex information, and the sex ratios within each of the classification schemes are as follows: SP10.1 202 F/199 M, SP10.2 202/199, SP10.3 202/199, SP14.1 177/178, SP14.2 136/125, Morph 112/120.

@@

% predictors/features/covariates
%   1:25 PCs
%   size
%   size X PC1
% analogy to PCA regression
The maximum set of possible predictors or features used for any model of our dataset is comprised of the first 25 principal components (PCs), scaled centroid size, the interaction between scaled centroid size and PC 1, and the interaction between scaled centroid size and PC 2. Additional interaction terms were not considered because of model complexity/sample size concerns. Size and the interaction between size and PC 1 were included as predictors to account for known ontogenetic variation in plastron shape \cite{Angielczyk2013a} as well as potential size differences between classes, even if this is unlikely \cite{Seeliger1945,Holland1992}. These data constitute a ``maximum set'' because the best or selected models based on five-fold cross-validation need not, and likely will not, include all predictors possible (see below). Because our supervised learning models use PCs as predictors, this approach is in many ways analogous to PCA regression. PCA regression takes advantage of reduction and orthogonality PCs to improve regression fit \cite{Hastie2009}. Because the PCs of shape are by definition orthogonal, they can easily serve as independent predictors or features of class membership without fear of collinearity.

@@

% auc of roc as performance measure
%   relationship between false positive rate and true positive rate
%   varies between 0.5 and 1 (random -- perfect)
%   better performance than accuracy when groups unbalanced
In classification studies, such as this one, a common metric of performance is the receiver operating characteristic (ROC) which is the relationship between the false and true positive rates \cite{Hastie2009}. The area under the ROC curve (AUC) is the derived estimate of the model performance; AUC ranges from 0.5 to 1 which correspond to performance similar to random guesses and perfect classification rates, respectively \cite{Hastie2009}. Both ROC and AUC are preferable to simple classification accuracy when class membership is unbalanced, as it is in these analyses \cite{Hastie2009}. The standard ROC and AUC calculations are defined only for binary classifications, which is not the case for our eight species and \textit{Emys} complex datasets. To generalize this approach for situations with multiple response classes, we used an all-against-one strategy where the model AUC is the average of the AUC values from the multiple binary comparisons of one class compared to all others \cite{Hand2001}.

@@

The results of the PCA of plastron shape in the \textit{Emys marmorata} dataset show no clear connection between plastron shape and any of the proposed classification schemes (Fig. \ref{fig:emys_pca}). The first PC axis of shape variation appears to be primarily structured by differences in individual centroid size (Fig. \ref{fig:emys_pca}); this was the motivation for including centroid size and its interaction with PC1 as predictors in all of the supervised learning models.

@@

The results from the eight species and \textit{Trachemys} datasets indicate that both of these classification schemes are more recognizable than not given our estimate of digitization error (Table \ref{tab:rep_res}). In contrast, the different \textit{E. marmorata} classification schemes appear to barely be distinct, with their within:between ratios approximating 1. This indicates that the magnitude of the differences between groupings is approximately the same as the difference between any two random individuals (Table \ref{tab:rep_res}).

@@

Analysis of the eight morphologically- and genetically-distinct species and the \textit{T. scripta scripta}--\textit{T. scripta elegans} datasets indicate that these taxa are sufficiently morphologically distinct to be differentiated on the basis of plastron shape. Both in-sample and out-of-sample classification have AUC values of approximately 1 for all methods, implying near-perfect classification rates (Fig. \ref{fig:other_sel}, \ref{fig:other_oos}). For both datasets, the ROC scores from testing datasets are tightly clustered near AUC = 1 (Fig. \ref{fig:other_oos}). These results demonstrate that when there are distinctions between the states of the classification schemes (i.e., differences in plastron shape that correlate with the different taxonomic groups), the methods used here can recover them.

>>>>

% Title must be 250 characters or less.
\begin{flushleft}
  {\Large
  \textbf\newline{Ensemble approaches for estimating congruence between species delimitation and morphological variation: a case study using the Pacific Pond Turtle (\textit{Emys marmorata})} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
  }
  \newline
  % Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
  \\
  Peter D Smits\textsuperscript{1*},
  Kenneth D Angielczyk\textsuperscript{2},
  Bryan L Stuart\textsuperscript{3},
  James F Parham\textsuperscript{4},
  \\
  \bigskip
  \textbf{1} Department of Integrative Biology, University of California -- Berkeley, Berkeley, California, USA.
  \\
  \textbf{2} Integrative Research Center, Field Museum of Natural History, Chicago, Illinois, USA.
  \\
  \textbf{3} Section of Research and Collections, North Carolina Museum of Natural Sciences, Raleigh, North Carolina, USA.
  \\
  \textbf{4} John D. Cooper Archaeological and Paleontological Center, Department of Geological Sciences, California State University -- Fullerton, Fullerton, California, USA.
  \\
  \bigskip

@@

\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
  We investigated the morphometric identification of cryptic species using machine learning approaches by examining their implications for a recently proposed cryptic turtle species (\textit{Emys pallida}). We used landmark-based morphometric data to quantify plastron shape in 532 adult \textit{E. marmorata/``E. pallida''} museum specimens. We assigned a classification to each specimen for six different binning schemes based on geographic occurrence data recorded in museum collection archives. We used an ensemble of supervised machine learning approaches to determine which classification hypothesis was best supported by the data. In addition, we applied the same approach to two clear-cut examples, one consisting of seven unambiguously distinct species closely related to \textit{E. marmorata} and an outgroup (\textit{Chrysemys picta}), and the other consisting of two subspecies of \textit{Trachemys scripta}. The analyses of the clear-cut examples produced near perfect classifications, demonstrating that plastron shape typically is a useful marker for differentiating turtle species, and that the methods can recover correct results when an appropriate signa exists. There is no clear ``best'' grouping of \textit{E. marmorata/``E. pallida''} based on plastron shape. Explanations for the lack of grouping in \textit{E. marmorata} include the possibility that genetic differentiation is not associated with plastron shape variation below the species level and/or that local selective pressures (e.g., from hydrological regime) overwhelm morphological differentiation. A reconsideration of the methods used to delimit \textit{``E. pallida,''} the lack of barriers to gene flow, the strong evidence for widespread admixture between lineages, and the fact that plastron shape can be used to delineate other emydid species and sub-species suggest that its lack of diagnosability most likely reflects the non-distinctiveness of this proposed taxon.

@@

These considerations have sparked interest in whether geometric morphometric analyses can capture fine-scale variation that can be used for identifying cryptic species. Most studies focus on using morphometrics to discover differences between taxa that were identified by other means \cite{Polly2003,Zelditch2004,Gaubert2005b,Gunduz2007,Polly2007a,Demandt2009,Markolf2013,Fruciano2016}. Additionally, there has been work on automated taxon identification and classification of taxa into groups \cite{Baylac2003,Dobigny2003,MacLeod2007,VandenBrink2011,Vitek2017}, as well as the development of models that combine genetic, phenotypic, and geographic data to infer evolutionary units of interest \cite{Guillot2012}.

Here, we investigate the morphometric identification of cryptic species using machine learning approaches. We use an ensemble learning approach where multiple methods are used in order to look for consensus among their results. We test our approach on three datasets: plastron (ventral shell) shape of seven species of closely related turtles and one outgroup, plastron shape of two subspecies of a single turtle species, and plastron shape of the \textit{Emys marmorata} species complex. In particular, we ask whether it is possible to determine which among a set of classification hypotheses best aligns with the observed morphology, and examine the implications of our results for the \textit{E. marmorata} complex.

% past work on automatic taxon identification and older approaches to classifying taxa
% why use machine learning methods
\subsection*{Background and study system}
Machine learning is an extension of known statistical methodology that emphasizes predictive accuracy and generality, often at the expense of the interpretability of individual parameters \cite{Hastie2009}. Basic statistical approaches are supplemented by randomization, sorting, and partitioning algorithms, along with the maximization or minimization of summary statistics, in order to best estimate a general model for all data, both sampled and unsampled \cite{Hastie2009}. Machine learning approaches have found use in medical research, epidemiology, economics, and automated identification of images such as handwritten zip codes \cite{Hastie2009}. % additional/more specific CITATIONS?

There are two major classes of machine learning methods: unsupervised and supervised learning. Unsupervised learning methods are used with unlabeled data where the underlying structure is estimated; they are analogous to clustering and density estimation methods \cite{Kaufman1990}. Supervised learning methods are used with labeled data where the final output of data is known and the rules for going from input to output are inferred. These are analogous to classification and regression models \cite{Breiman1984,Hastie2009}. Our application of the supervised learning approaches used in this study illustrates only a sampling of the various methods available for fitting classification models. The specific methods we used were chosen because they are suited for cases with more two or more response classes.

@@

More recently, the phylogeography of \textit{E. marmorata} and the possibility of cryptic diversity was investigated using molecular data \cite{Spinks2005,Spinks2010,Spinks2014}. Based on mitochondrial DNA, Spinks and Shaffer \cite{Spinks2005} recognized four subclades within \textit{E. marmorata}, a northern clade, a San Joaquin Valley clade, a Santa Barbara clade, and a southern clade. Analyses with nuclear DNA \cite{Spinks2010} and single-nucleotide polymorphism (SNP) data suggest a primarily north--south division in \textit{E. marmorata}, although these datasets differed from the mitocondrial-based results of Spinks and Shaffer \cite{Spinks2005} in the location of the break point \cite{Spinks2014}. All three studies discussed the potential taxonomic implications of their results, with Spinks et al. \cite{Spinks2014} going so far as to strongly advocate for the recognition of at least two species (\emph{E. marmorata} and \emph{E. pallida}), and a possible third based on populations in Baja California. However, they did not discuss in detail the morphological characters that would help to diagnose these species beyond those specified by Seeliger \cite{Seeliger1945}. Given that these characters are variable within the proposed species, and that Holland \cite{Holland1992} described shell shape variation that might be consistent with this taxonomy, a geometric morphometric analysis of shell shape might provide a reliable way to diagnose groups (whether species or subspecies) within \textit{E. marmorata}.

@@

We choose to analyze plastron shape for multiple reasons. First, it is very easy to collect geometric morphometric data on plastron shape from two-dimensional photographs as the structure is virtually flat. This approach allows both museum specimens and individuals photographed in the field to be analyzed together. Second, previous work has suggested that there are strong differences in plastron shape among traditionally-recognized emydine species \cite{Angielczyk2007,Angielczyk2011,Angielczyk2013a}. Finally, due to these previous studies a large dataset was readily available.

@@

Three different geometric morphometric datasets describing turtle plastron variation were assembled for this analysis: 1) specimens from seven distinct emydine species and a deriochelyine outgroup (\textit{C. picta}); 2) \textit{T. scripta} specimens from the two main subspecies (\textit{T. scripta elegans} and \textit{T. scripta scripta}); and 3) \textit{E. marmorata} specimens from across the species' geographic range. The first two datasets are intended to serve as a test of whether machine learning techniques can differentiate species-level groupings of emydid turtles using plastron shape. We expect that the first case represents a low complexity dataset because of the high level of plastron shape disparity that exists among these species \cite{Claude2003a,Claude2006,Angielczyk2011}, whereas the second dataset should be relatively higher in complexity and more analogous to the \textit{E. marmorata} example. We predict that the \textit{E. marmorata} dataset should be of the highest complexity and our greatest challenge given the previous finding that only very subtle differences existed between geographically-distinct populations \cite{Holland1992}.
%We expect the first dataset represents a low-complexity dataset given the high level of plastron shape disparty that exists among these species \cite{Angielczyk2011}, whereas the second dataset should be relatively higher in complexity than the first. Ultimately, we expect that the third dataset to be of the highest complexity and our biggest analytical challenge. For all datasets we chose to focus on adults because significant changes in plastron shape occur over the course of ontogeny in \textit{E. marmorata} and other emydines \cite{Angielczyk2013a}.
% three turtle datasets
%   eight species
%   Trachemys species
%   Emys complex
% 26 landmarks

@@

The final dataset is of 532 adult \textit{E. marmorata} museum specimens, though not all specimens were able to be assigned a class for all schemes (Fig. \ref{fig:map}). These specimens represent a subset of those included in previous studies \cite{Angielczyk2007,Angielczyk2011,Angielczyk2013a}. Because previous genetic studies did not use vouchered specimens \cite{Spinks2005,Spinks2010,Spinks2014} we were not able to directly sample the individuals in their studies. Instead, our specimen classifications were based solely on the geographic information and not explicit assignment using molecular data. For each taxonomic hypothesis, specimens were assigned to one of the possible classes based on geographic occurrence data recorded in museum collections. In cases where precise latitude and longitude information were not available we estimated them from other locality information. Because the exact barriers between different biogeographic regions are unknown and unclear, we represented each hypothesis with multiple possible realizations representing the classification uncertainty for specimens present at the geographic boundaries. The taxonomic hypotheses and sub-hypotheses for \textit{E. marmorata} used here are presented in Table \ref{tab:hypotheses} and Fig. \ref{fig:map}.

We used three main binning schemes. All three schemes include a class for \textit{E. marmorata} specimens from northern populations (marm) as well as a class for those assigned to \textit{E. pallida} (pall) and an intergrade zone in the Central Coast Ranges (CCR). The schemes differ in the assignment of samples from the San Joaquin Valley (Fig. \ref{fig:map}). Scheme SP10.1 and SP10.2 differ in the assignment of specimens from the western San Joaquin Valley to either CCR or marm reflecting uncertainty regarding their genetic affinity as explained above. In scheme SP10.3 these specimens are assigned to a San Joaquin class reflecting the mitochondrial distinctiveness shown by Spinks and Shaffer \cite{Spinks2005}. We used two versions of the binning scheme based on the results of Spinks et al. \cite{Spinks2014}: SP14.1 reflects their phylogenetic network analysis and SP14.2 corresponds to their Bayesian species delimitation analysis. SP14.2 requires the addition of two new classes, ``Baja'' and ``Foothill,'' to accommodate the genetic groupings recovered by the SNP Structure analysis that was used to create the guide tree for the BPP species delimitation analysis \cite{Spinks2014}. Finally, we proposed a conservative morphological scheme (``Morph'') in order to compare the molecular hypotheses with something approximating the original taxonomic hypothesis for the group. This scheme is made up solely of the marm and pall classes from the SP10.3 scheme.

@@

Sex was known only for a subset of the total dataset and was not included as a predictor of classification. Instead, we estimated the degree to which specimens cluster morphologically by sex in order to determine how much of a potential biasing factor sexual dimorphism could be for our analysis of the \textit{E. marmorata} species complex (see below).
%The scheme names are as follows: Mito 1 and 2 correspond to \cite{Spinks2005}, Mito 3 corresponds to \cite{Spinks2010}, Morph 1 and Morph 2 correspond to \cite{Holland1992}, and Nuclear corresponds to \cite{Spinks2014}. % CHANGE

Following previous work on plastron shape \cite{Angielczyk2007,Angielczyk2011,Angielczyk2013a}, we used TpsDig 2.04 \cite{Rohlf2005} to digitize 19 two-dimensional landmarks (Fig. \ref{fig:plastra}). Seventeen of the landmarks are at the endpoints or intersection of the keratinous scutes that cover the plastron. Twelve of the landmarks were symmetrical across the axis of symmetry. Because damage prevented the digitization of all the symmetric landmarks in some specimens, we reflected landmarks across the axis of symmetry (i.e. midline) prior to analysis and used the average position of each symmetric pair. In cases where damage or incompleteness prevented symmetric landmarks from being determined, we used only the single member of the pair. We conducted all subsequent analyses on the resulting ``half'' plastra. We superimposed the plastral landmark configurations using generalized Procrustes analysis \cite{Dryden1998a}, after which we calculated the principal components (PC) of shape using the \texttt{shapes} package for R \cite{R2016,Dryden2013}. All specimens were used for superimposition, after which the subset labeled for each of the schemes were used in model training and testing (see below).

\subsection*{Biasing effects}
% digitization
We estimated the possible effect of digitization error \cite{Arnqvist1998,Cramon2007,Munoz-MunozF.2010} on our results by comparing within-specimen (replicated) Procrustes distances to the distances between classification scheme centroids. Ten randomly-selected \textit{E. marmorata} specimens were each digitized four times, with the original set of digitized coordinates serving as a fifth replicate. These 50 landmark configurations were Procrustes superimposed. A range of four Procrustes distances was then calculated as the average of the pairwise distances between each of the replicate configurations of a given specimen.

For each specimen, the difference in shape caused by digitization was calculated as the mean of all pairwise Procrustes distances between the five replicates of that specimen. The average distance between any two digitizations was calculated as the mean of all pairwise Procrustes distances between all replicates for all specimens. The ratio between these two values was used to assess the magnitude of variation caused by digitization. The goal of this ratio is to determine if the within group distances are smaller on average than the between individual distances; a value of 0 indicates perfect grouping, a value of 1 indicates no difference between grouping and no grouping, and a value of 1+ indicates that the grouping is counter-intuitive to the data.

\textit{Emys marmorata} is known to display sexual dimorphism in plastral shape, particularly the presence of a plastral concavity in males \cite{Seeliger1945}. To test for biases resulting from sexual dimorphism in our \textit{E. marmorata} dataset, we used a simple permutation test to determine if the distance between the mean female and male shapes is greater than expected when the sex labels are randomly shuffled. Because not all of our specimens have sex identifications associated with them, this analysis was done using a subset of the data (257 of 532). This analysis was also done for each of the classification schemes in order to determine if sex is a biasing factor within these groups. As with the total dataset, only a limited number of specimens in each classification scheme have sex information, and the observed sex ratios for each of the classes within the classification schemes are summarized in Table \ref{tab:sex_ratio}. Importantly, these within-class sex ratios are based only on the subset of specimens with sex information.

\begin{table}
  \centering
  \caption{Table of within-class sex ratios from those speciems with sex information for ecah classification schemes used in this analysis.}
    \begin{tabular}{l l}
      \hline
      Scheme & class ratios \\
      \hline
      SP10.1 & CCR: 70 F/60 M; Marm: 112/113; Pall: 20/26 \\
      SP10.2 &  CCR: 63/50; Marm: 119/123; Pall: 20/26 \\
      SP10.3 & CCR: 63/50; Marm: 92/95; Pall: 20/26; SJ: 27/28 \\
      SP14.1 & Marm: 112/117; Pall: 65/61 \\
      SP14.2 & Baja: 0/5; Foothill: 20/19; Marm: 70/63; Pall: 45/38\\
      Morph & Marm: 92/94; Pall: 20/26 \\
      \hline
    \end{tabular}
  \label{tab:sex_ratio}
\end{table}

@@

% predictors/features/covariates
%   1:25 PCs
%   size
%   size X PC1
% analogy to PCA regression
The maximum set of possible predictors or features used for any model of our dataset is comprised of the first 25 principal components (PCs), scaled centroid size, the interaction between scaled centroid size and PC 1, and the interaction between scaled centroid size and PC 2. Additional interaction terms were not considered because of model complexity/sample size concerns. Size and the interaction between size and PCs 1 and 2 were included as predictors to account for known ontogenetic variation in plastron shape \cite{Angielczyk2013a} as well as potential size differences between classes, even if this is unlikely \cite{Seeliger1945,Holland1992}. These data constitute a ``maximum set'' because the best or selected models based on five-fold cross-validation need not, and likely will not, include all predictors possible (see below). Because our supervised learning models use PCs as predictors, this approach is in many ways analogous to PCA regression. PCA regression takes advantage of the reduction and orthogonality PCs to improve regression fit \cite{Hastie2009}. Because the PCs of shape are by definition orthogonal, they can easily serve as independent predictors or features of class membership without fear of collinearity.

@@

% auc of roc as performance measure
%   relationship between false positive rate and true positive rate
%   varies between 0.5 and 1 (random -- perfect)
%   better performance than accuracy when groups unbalanced
In classification studies, such as this one, a common metric of performance is the receiver operating characteristic (ROC) which is the relationship between the false and true positive rates \cite{Hastie2009}. The area under the ROC curve (AUC) is the derived estimate of model performance; AUC ranges from 0.5 to 1 which correspond to performance similar to random guesses and perfect classification rates, respectively \cite{Hastie2009}. Both ROC and AUC are preferable to simple classification accuracy when class membership is unbalanced, as it is in these analyses \cite{Hastie2009}. The standard ROC and AUC calculations are defined only for binary classifications, which is not the case for our eight species and \textit{Emys} complex datasets. To generalize this approach for situations with multiple response classes, we used an all-against-one strategy where the model AUC is the average of the AUC values from the multiple binary comparisons of one class compared to all others \cite{Hand2001}.

@@

The results of the PCA of plastron shape in the \textit{Emys marmorata} dataset show no clear connection between plastron shape and any of the proposed classification schemes (Fig. \ref{fig:emys_pca}). The first PC axis of shape variation appears to be primarily structured by differences in individual centroid size (Fig. \ref{fig:emys_pca}); a linear regression of PC 1 vs log centroid size shows that size explains approximately 81 percent of the variation along this axis. This was the motivation for including centroid size and its interaction with PC1 as predictors in all of the supervised learning models. Size typically explains less than three percent of the variance of the remaining PC axes (e.g. 2.5\% on PC 2).

@@

The results from the eight species and \textit{Trachemys} datasets indicate that both of these classification schemes are more recognizable than not given our estimate of digitization error (Table \ref{tab:rep_res}). In contrast, the different \textit{E. marmorata} classification schemes appear to be barely distinct, with their within:between ratios approximating 1. This indicates that the magnitude of the differences between groupings is approximately the same as the difference between any two random individuals (Table \ref{tab:rep_res}).

@@

Analysis of the eight morphologically- and genetically-distinct species and the \textit{T. scripta scripta}--\textit{T. scripta elegans} datasets indicate that these taxa are sufficiently morphologically distinct to be differentiated on the basis of plastron shape. Both in-sample and out-of-sample classifications have AUC values of approximately 1 for all methods, implying near-perfect classification rates (Fig. \ref{fig:other_sel}, \ref{fig:other_oos}). For both datasets, the ROC scores from testing datasets are tightly clustered near AUC = 1 (Fig. \ref{fig:other_oos}). These results demonstrate that when there are distinctions between the states of the classification schemes (i.e., differences in plastron shape that correlate with the different taxonomic groups), the methods used here can recover them.
